Code Club Project: Assessing whether intra and inter-genomic variation hinder utility of ESVs and ASVs

### 2021-01-20 submission/manuscript.tex
Generate lumping and splitting rate figure
Generate ROC curve figure
Create figure captions and integrate numbers into manuscript

To investigate the variation in the number of copies of the 16S rRNA
gene per genome as well as the intragenomic variation among copies of
the 16S rRNA gene, I obtained reference 16S rRNA sequences from the rrn
copy number database (rrnDB; CITATION). Among the \textbf{4,774} species
represented in the rrnDB there were \textbf{15,614} genomes. The median
number of rrn operson per species ranged between \textbf{1}
(e.g.~\textbf{\emph{Mycobacterium tuberculosis}}) and \textbf{19}
(\textbf{\emph{Metabacillus litoralis}}) copies of the rrn operon. As
the number of copies of the operon in a genome increased, the number of
variants of the 16S rRNA gene in each genome also increased
(\textbf{FIGURE}). On average, there were \textbf{1} variants per copy
of the full length 16S rRNA gene and an average of \textbf{0.26},
\textbf{0.33}, and \textbf{0.27} variants when considering the V4,
V3-V4, and V4-V5 regions of the gene, respectively. Although a species
tended to have a consistent number of 16S rRNA gene copies per genome,
the number of total variants increased with the number of genomes that
were sampled (\textbf{Figure 1}). For example,
\textbf{\emph{Mycobacterium tuberculosis}} generally only had \textbf{1}
copy of the gene per genome, but across the \textbf{180} genomes that
have been sequenced there were \textbf{11} versions of the gene.
Similarly, a \emph{E. coli} genome typically had \textbf{7} copies of
the 16S rRNA gene with between \textbf{6} and \textbf{10} distinct full
length sequences per genome. Across the \textbf{958} \emph{E. coli}
genomes that have been sequenced, there were \textbf{1,013} different
variants of the gene. These observations highlight the risk of selecting
a threshold for defining units of inference that is too narrow because
it is possible to split a single genome into multiple units.

A method to avoid splitting a single genome into multiple units of
inference is to cluster 16S rRNA gene sequences together that are
similar. However, this also increases the risk of lumping together genes
from different species that are similar to each other. Therefore, I
assessed the impact of the threshold used to define clusters of 16S rRNA
genes on the propensity to split a genome apart or to lump species
together. For full length 16S rRNA gene sequences, I found that at a
threshold of \textbf{5.5}\%, 95\% of the species with 7 copies of the
rrn operon would be represented by a single OTU. Similarly, thresholds
of \textbf{2.5}, \textbf{4.0}, and \textbf{3.5}\% were observed for the
V4, V3-V4, and V4-V5 regions, respectively. However, at these
thresholds, multiple species could be represented by the same OTU. At
the highest level of resolution, \textbf{3.6}\% of the species shared a
16S rRNA gene sequence variant with another species when considering
full length sequences and \textbf{14.9}, \textbf{10.2}, and
\textbf{12.0}\% when considering the V4, V3-V4, and V4-V5 regions,
respectively. At the commonly used 3\% threshold, \textbf{25.2}\% of the
species shared an OTU when considering full length sequences and
\textbf{33.0}, \textbf{29.4}, and \textbf{32.2}\% when considering the
V4, V3-V4, and V4-V5 regions, respectively. Given the risk of splitting
a genome into multiple OTUs is more biologically problematic than
lumping species together, larger thresholds are advisable.

To provide a more nuanced approach to selecting a threshold, it would be
useful to to quantify the sensitivity and specificity of characterizing
bacterial species using OTUs defined at different thresholds. I created
confusion matrices for multiple regions of the 16S rRNA gene: true
positives were those cases where two ASVs were joined in the same OTU
and the same species; true negatives were those cases where two ASVs
from different OTUs came from different species; false positives were
those ASVs that joined the same OTU, but were from different species;
and false negatives were those ASVs that joined different OTUs, but were
from the same species. By calculating the sensitivity and specificity
for each threshold and each region of the 16S rRNA gene, I was able to
constuct a receiver operator characteristic curve (ROC). Because the ROC
curve represents a range of possible thresholds and sensitivities and
specificities, I used two metrics to select the best threshold for
defining an OTU. First, I identified the thresholds where the
sensitivity and specificity were most similar to each other. For this
criterion, the best distance thresholds were \textbf{6.0}\% (V1-V9),
\textbf{4.5}\% (V4), \textbf{5.5}\% (V3-V4), and \textbf{4.0}\% (V4-V5).
Second, I identified the distance threshold that resulted in the point
on the ROC curve that was closest to perfect classification. For this
criterion, the best distance thresholds were \textbf{5.5}\% (V1-V9),
\textbf{3.5}\% (V4), \textbf{4.5}\% (V3-V4), and \textbf{3.5}\% (V4-V5).
Surprisingly, these analyses revealed that thresholds near 3\% distance
balance the risks of splitting genomes into separate OTUs and lumping
species into the same OTU.

@@

\textbf{Figure 1. ESV rate increases as the number of genomes sampled
per species increases}

>>>>

To investigate the variation in the number of copies of the 16S rRNA
gene per genome as well as the intragenomic variation among copies of
the 16S rRNA gene, I obtained reference 16S rRNA sequences from the rrn
copy number database (rrnDB; CITATION). Among the \textbf{4,774} species
represented in the rrnDB there were \textbf{15,614} genomes. The median
number of rrn operson per species ranged between \textbf{1}
(e.g.~\textbf{\emph{Mycobacterium tuberculosis}}) and \textbf{19}
(\textbf{\emph{Metabacillus litoralis}}) copies of the rrn operon. As
the number of copies of the operon in a genome increased, the number of
variants of the 16S rRNA gene in each genome also increased. On average,
there were \textbf{1} variants per copy of the full length 16S rRNA gene
and an average of \textbf{0.26}, \textbf{0.33}, and \textbf{0.27}
variants when considering the V4, V3-V4, and V4-V5 regions of the gene,
respectively. Although a species tended to have a consistent number of
16S rRNA gene copies per genome, the number of total variants increased
with the number of genomes that were sampled (\textbf{Figure 1}). For
example, \textbf{\emph{Mycobacterium tuberculosis}} generally only had
\textbf{1} copy of the gene per genome, but across the \textbf{180}
genomes that have been sequenced there were \textbf{11} versions of the
gene. Similarly, a \emph{E. coli} genome typically had \textbf{7} copies
of the 16S rRNA gene with between \textbf{6} and \textbf{10} distinct
full length sequences per genome. Across the \textbf{958} \emph{E. coli}
genomes that have been sequenced, there were \textbf{1,013} different
variants of the gene. These observations highlight the risk of selecting
a threshold for defining units of inference that is too narrow because
it is possible to split a single genome into multiple units.

A method to avoid splitting a single genome into multiple units of
inference is to cluster 16S rRNA gene sequences together that are
similar. However, this also increases the risk of lumping together genes
from different species that are similar to each other. Therefore, I
assessed the impact of the threshold used to define clusters of 16S rRNA
genes on the propensity to split a genome apart or to lump species
together. For full length 16S rRNA gene sequences, I found that at a
threshold of \textbf{5.5}\%, 95\% of the species with 7 copies of the
rrn operon would be represented by a single OTU. Similarly, thresholds
of \textbf{2.5}, \textbf{4.0}, and \textbf{3.5}\% were observed for the
V4, V3-V4, and V4-V5 regions, respectively. However, at these
thresholds, multiple species could be represented by the same OTU
(\textbf{Figure 2}). At the highest level of resolution, \textbf{3.6}\%
of the species shared a 16S rRNA gene sequence variant with another
species when considering full length sequences and \textbf{14.9},
\textbf{10.2}, and \textbf{12.0}\% when considering the V4, V3-V4, and
V4-V5 regions, respectively. At the commonly used 3\% threshold,
\textbf{25.2}\% of the species shared an OTU when considering full
length sequences and \textbf{33.0}, \textbf{29.4}, and \textbf{32.2}\%
when considering the V4, V3-V4, and V4-V5 regions, respectively. Given
the risk of splitting a genome into multiple OTUs is more biologically
problematic than lumping species together, larger thresholds are
advisable.

To provide a more nuanced approach to selecting a threshold, it would be
useful to to quantify the sensitivity and specificity of characterizing
bacterial species using OTUs defined at different thresholds. I created
confusion matrices for multiple regions of the 16S rRNA gene: true
positives were those cases where two ASVs were joined in the same OTU
and the same species; true negatives were those cases where two ASVs
from different OTUs came from different species; false positives were
those ASVs that joined the same OTU, but were from different species;
and false negatives were those ASVs that joined different OTUs, but were
from the same species. By calculating the sensitivity and specificity
for each threshold and each region of the 16S rRNA gene, I was able to
constuct a receiver operator characteristic curve (ROC; \textbf{Figure
S1}). Because the ROC curve represents a range of possible thresholds
and sensitivities and specificities, I used two metrics to select the
best threshold for defining an OTU. First, I identified the thresholds
where the sensitivity and specificity were most similar to each other.
For this criterion, the best distance thresholds were \textbf{6.0}\%
(V1-V9), \textbf{4.5}\% (V4), \textbf{5.5}\% (V3-V4), and \textbf{4.0}\%
(V4-V5). Second, I identified the distance threshold that resulted in
the point on the ROC curve that was closest to perfect classification.
For this criterion, the best distance thresholds were \textbf{5.5}\%
(V1-V9), \textbf{3.5}\% (V4), \textbf{4.5}\% (V3-V4), and \textbf{3.5}\%
(V4-V5). Surprisingly, these analyses revealed that thresholds near 3\%
distance balance the risks of splitting genomes into separate OTUs and
lumping species into the same OTU.

@@

\textbf{Figure 1. The ratio of number of distinct ASVs per copy of the
\emph{rrn} operon increases for a species as the number of genomes
sampled increases.} Each point represents a different species and is
shaded to be 80\% transparent so that when points overlap they become
darker. The blue line represents a smoothed fit through the data.

\includegraphics{../figures/lump_split.pdf}

\textbf{Figure 2. As the distance threshold used to define an OTU
increases, the fraction of genomes split into separate OTUs decreases
while the fraction of species that are merged into the same OTU
increases.} These data represent the median fractions for both
measurements across 100 randomizations. In each randomization, one
genome was sampled from each species.

\includegraphics{../figures/roc_curve.pdf}

\textbf{Figure S1. Distance thresholds larger than 3\% provide better
sensitivity and specificity when assigning ASVs to OTUs when trying to
represent species-level classifications.} The sensitivities and
specificities for ASVs were \textbf{0.435/1.000}, \textbf{0.748/1.000},
\textbf{0.887/0.999}, and \textbf{0.853/0.999} for the V1-V9, V3-V4, V4,
V4-V5 regions, respectively. The gray diagonal line represents the
position of where the sensitivity and specificity were equal. These data
represent the median fractions for sensitivity and specificity across
100 randomizations. In each randomization, one genome was sampled from
each species.
